---
title: "STAT 6440: Data Mining: Group 8"
author: "Final source code"
subtitle: "Group Assignment 3"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("myfunctions.R")
library(caret)
library(car)
library(FNN)
library(rpart)
library(rpart.plot)
library(gains)
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(skimr)
library(ggcorrplot)
library(fastAdaboost)
```



### Exploratory Analysis


In the part of our we want to explore the dataset and understand the represenatation of each features. 


```{r,echo=TRUE}
# load the dataset
dataLifeExpectancy <- read.csv("Life-Expectancy-Data-Updated.csv")
str(dataLifeExpectancy)
print(dim(dataLifeExpectancy))
```

\textbf{Data characteristics}

In we want to understand the information each feature in the data set and capture missing information in data. From the data, we observed the data was cleaned and there were no missing values. The total number of observation collected are 2864 and 21 individual features including the response variable. 

```{r,echo=TRUE}
colSums(is.na(dataLifeExpectancy))

head(dataLifeExpectancy)
```

- This data were collected from various continents in the world showing the occurence of death. The table shows the number of observations collected from each continent. From the data we observed that majority of observations were collected from Africa representing 816 and least collected is North America with an amount of 48. 

```{r,echo=TRUE}
table(dataLifeExpectancy$Region)

# extract the features of interest 
life_expectancy_num_vars <- dataLifeExpectancy %>% 
  select(where(is.numeric)) %>%
  select(-Year)
```

- We want to have the summary statistic of country and their life expectancy. Extracted few of the features to under summary information under each continents using ; mortality information both infants and adult. 

From the data we observed that Africa has the lowest median life expectancy among all the continent with a value of $57.05$ follow by Oceania and Asia.  

- Interestingly, Africa has highest infants mortality follow by the Asia continent

- We only have to two categorical features ; Economy_status_Developed and Economy_status_Developing. This was generated based the Status from the original datasets. 

```{r,echo=TRUE}
summary(life_expectancy_num_vars)
```





- Distribution of each numeric features. We wanted to understand the skewness of the distribution under each feature of the dataset. We observed that most of the features are righted and lefted skewed in nature 

```{r,echo=TRUE}
# extract the features of interest 
num_vars <- dataLifeExpectancy %>% 
  select(where(is.numeric)) %>%
  select(-Year)

# All histograms at once
num_vars %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free", ncol = 4) +
  labs(title = "Distributions of All Numeric Features") +
  theme_minimal(base_size = 9)
```


In this part of our analysis, we want to understand the distribution of the life expectancy by region. We observed some level of outlying points for some of the features using box and whiskey plot. 


```{r,echo=TRUE}
# 
ggplot(dataLifeExpectancy, aes(x = Region, y = Life_expectancy, fill = Region)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 16, alpha = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 23,
               size = 3, fill = "white", color = "black") +  # mean diamond
  labs(
    title    = "Distribution of Life Expectancy by Region",
    x        = "Region",
    y        = "Life Expectancy",
    fill     = "Region"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5)) +
  theme(legend.position = "none")

```

- Individual distribution without considering the region. 

```{r,echo=TRUE}
num_vars %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Variable, y = Value)) +
  geom_boxplot(fill = "steelblue", outlier.colour = "red",
               outlier.size = 1, alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free", ncol = 4) +
  labs(title = "Outlier Detection Across Numeric Variables") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_blank())
```

- This is trend of average life expectancy over the years combining all information from the regions. We wanted to group by the years and under the trends of life expectancy over years. We observed that there is an increase in life expectancy over years combined information. 


```{r,echo=TRUE}
dataLifeExpectancy %>%
  group_by(Year) %>%
  summarise(mean_le = mean(Life_expectancy, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_le)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(size = 2) +
  labs(title = "Average Life Expectancy Over Time",
       x = "Year", y = "Mean Life Expectancy") +
  theme_minimal()
```


```{r,echo=TRUE}
# extract the numeric information 
num_vars <- dataLifeExpectancy %>% 
  select(where(is.numeric)) %>%
  select(-Year)

cor_matrix <- cor(num_vars, use = "pairwise.complete.obs")

# Correlation with target only
cor_with_target <- sort(cor_matrix[, "Life_expectancy"], decreasing = TRUE)
print(cor_with_target)

# Full heatmap
ggcorrplot(round(cor_matrix,1),
           method = "square",
           type   = "upper",
           lab    = TRUE,
           show.diag = TRUE,
           lab_size = 1.9,
           colors = c("#d73027", "white", "#1a9850"),
           title  = "Correlation Matrix") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7))
```

```{r,echo=TRUE}
library(kableExtra)

cor_df <- as.data.frame(round(cor_matrix, 2))

cor_df %>%
  kbl(caption = "Correlation Matrix — WHO Life Expectancy") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 11) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE, background = "#f5f5f5")
```


From the datasets, we observation the features were cleaned without missing values. 

### Ensemble models implementation 

In this session, we will extract all the numeric information and after we will split the data into train and test. We will fit with the train set and assess the performance of the best model selected. 

```{r,echo=TRUE}
# setting a seed 
RNGkind (sample.kind = "Rounding") 
set.seed(0) # set a seed 
p2_Expectancy <- partition.2(num_vars, 0.8)
training_dataExpectancy <- p2_Expectancy$data.train
test_dataExpectancy <- p2_Expectancy$data.test

cat("Dimension of the train set: ", dim(training_dataExpectancy),"\n")
cat("Dimension of the test set: ",dim(test_dataExpectancy))
```

We will implement the model implementation pipeline. 

```{r,echo=TRUE}
# set the cross-validation
ctrl <- trainControl(
  method    = "cv",
  number    = 5,          # 5-fold CV
  verboseIter = FALSE,
  allowParallel = TRUE
)
```

#### Bagging 

In this case, we perform 5-folds cross-validation 

```{r,echo=TRUE}
# time the bagging 
time_bag <- system.time({
  set.seed(123)
  model_bag <- train(
    Life_expectancy ~ .,
    data      = training_dataExpectancy,
    method    = "treebag",
    trControl = ctrl,
    nbagg     = 50
  )
})

cat("Bagging done\n")
print(model_bag)
print(time_bag)
```



#### Random Forest 


```{r,echo=TRUE}
time_rf <- system.time({
set.seed(123)
model_rf <- train(
  Life_expectancy ~ .,
  data      = training_dataExpectancy,
  method    = "rf",
  trControl = ctrl,
  tuneGrid  = expand.grid(mtry = c(3, 5, 7, 9)),
  ntree     = 500,
  importance = TRUE
)})

cat("Random Forest done\n")
print(model_rf)
plot(model_rf, main = "RF — mtry tuning")
print(time_rf)
```


#### Adaboost 

We had to perform hyperparameter tuning for number of boosting interactions, max tree depth and learning rate. 

```{r,echo=TRUE}
# AdaBoost in caret uses method = "adaboost"
# Requires: install.packages("fastAdaboost")
tuneGrid <- expand.grid(
    n.trees  = c(50, 100, 150),   # Number of boosting iterations
    interaction.depth = c(1, 2, 3), # Max tree depth (1 = stumps)
    shrinkage = c(0.1,0.2) ,# Learning rate (smaller = slower but more robust)
    n.minobsinnode    = c(10, 20)
  )

time_ada <- system.time({
set.seed(123)
modelLookup("gbm")
model_ada <- train(
  Life_expectancy ~ .,
  data      = training_dataExpectancy,
  method    = "gbm",
  trControl = ctrl,
  tuneGrid  = tuneGrid,
  verbose   = FALSE
)})

cat("AdaBoost done\n")
print(model_ada)
print(time_ada)

```

#### Stacking algorithms 

In we perform ensemble in different way other than the standard approach where we consider three different weaks models :

- Lasso 
- Ridge 
- Knn model 

We wanted to combine two shrinkage algorithms to assess how these model perform compare to tree ensemble. 



```{r,echo=TRUE}

# ----- tree regression ------
time_tree <- system.time({
  tune_tree <- train(
  Life_expectancy ~ .,
  data      = training_dataExpectancy,
  method    = "rpart",
  trControl = ctrl,
  tuneGrid  = expand.grid(cp = c(0.001, 0.005, 0.01, 0.05, 0.1))
)
})
# ---- Ridge ----
time_ridge <- system.time({
  set.seed(123)
  model_ridge <- train(
    Life_expectancy ~ .,
    data      = training_dataExpectancy,
    method    = "glmnet",
    trControl = ctrl,
    tuneGrid  = expand.grid(alpha = 0, lambda = 10^seq(-4, 2, length = 50)),
    preProcess = c("center", "scale")
  )
})

# ---- Lasso ----
time_lasso <- system.time({
  set.seed(123)
  model_lasso <- train(
    Life_expectancy ~ .,
    data      = training_dataExpectancy,
    method    = "glmnet",
    trControl = ctrl,
    tuneGrid  = expand.grid(alpha = 1, lambda = 10^seq(-4, 2, length = 50)),
    preProcess = c("center", "scale")
  )
})

# ---- KNN ----
time_knn <- system.time({
  set.seed(123)
  model_knn <- train(
    Life_expectancy ~ .,
    data      = training_dataExpectancy,
    method    = "knn",
    trControl = ctrl,
    tuneGrid  = expand.grid(k = c(3, 5, 7, 9, 11, 15)),
    preProcess = c("center", "scale")
  )
})



# 

# ---- Stack meta-learner ----
time_stack <- system.time({
  pred_ridge <- predict(model_ridge, test_dataExpectancy)
  pred_lasso <- predict(model_lasso, test_dataExpectancy)
  pred_knn   <- predict(model_knn,   test_dataExpectancy)
  pred_tree <- predict(tune_tree,test_dataExpectancy)
  meta_test  <- data.frame(
    ridge = pred_ridge,
    lasso = pred_lasso,
    knn   = pred_knn,
    tree = pred_tree,
    Life_expectancy = test_dataExpectancy$Life_expectancy
  )
  meta_model <- lm(Life_expectancy ~ ridge + lasso + knn + tree, data = meta_test)
  pred_stack <- predict(meta_model, meta_test)
})

cat("Result of the prediction= \n")
print(head(pred_stack))
print(head(test_dataExpectancy$Life_expectancy))
```


#### Evaluation of models 


```{r,echo=TRUE}

library(Metrics)

actual <- test_dataExpectancy$Life_expectancy

pred_bag   <- predict(model_bag$finalModel,   test_dataExpectancy)
pred_rf    <- predict(model_rf$finalModel,    test_dataExpectancy)
pred_ada   <- predict(model_ada$finalModel,   test_dataExpectancy)

get_metrics <- function(actual, predicted, model_name, elapsed_time) {
  tibble(
    Model        = model_name,
    RMSE         = round(rmse(actual, predicted), 4),
    MAE          = round(mae(actual, predicted),  4),
    Time_sec     = round(elapsed_time["elapsed"],  2)
  )
}

final_table <- bind_rows(
  get_metrics(actual, pred_bag,   "Bagging",                time_bag),
  get_metrics(actual, pred_rf,    "Random Forest",          time_rf),
  get_metrics(actual, pred_ada,   "AdaBoost",               time_ada),
  get_metrics(actual, pred_ridge, "Ridge",                  time_ridge),
  get_metrics(actual, pred_lasso, "Lasso",                  time_lasso),
  get_metrics(actual, pred_knn,   "KNN",                    time_knn),
  get_metrics(actual, pred_tree,   "tree",                    time_tree),
  get_metrics(actual, pred_stack, "Stack (Ridge+Lasso+KNN)", time_stack)
) %>% arrange(RMSE)

print(final_table)
```




#### Lift plot 


We want to evaluate each model using the lift plot to understand the rate of contribution based on the number of observation collected. 


- Bagging lift plot 


```{r,echo=TRUE}
# access the gain 
gain_bag <- gains(test_dataExpectancy$Life_expectancy, pred_bag) # creates the cumulative gain chart
print(gain_bag)

# Plot Lift chart: Percent cumulative response
x_bag <- c(0, gain_bag$depth)
pred.y_bag <- c(0, gain_bag$cume.pct.of.total)
avg.y_bag <- c(0, gain_bag$depth/100)
plot(x_bag, pred.y_bag, main = "Cumulative Lift Chart: bagging", xlab = "deciles", 
     ylab = "Percent cumulative response", type = "l", col = "blue", cex.lab = 1.5)
lines(x_bag, avg.y_bag, type = "l")
```


- Random Forest lift plot 


```{r,echo=TRUE}
# access the gain 
gain_rf <- gains(test_dataExpectancy$Life_expectancy, pred_rf) # creates the cumulative gain chart
print(gain_rf)

# Plot Lift chart: Percent cumulative response
x_rf <- c(0, gain_rf$depth)
pred.y_rf <- c(0, gain_rf$cume.pct.of.total)
avg.y_rf <- c(0, gain_rf$depth/100)
plot(x_rf, pred.y_rf, main = "Cumulative Lift Chart: Random forest", xlab = "deciles", 
     ylab = "Percent cumulative response", type = "l", col = "blue", cex.lab = 1.5)
lines(x_rf, avg.y_rf, type = "l")
```

- Adaboost lift plot


```{r,echo=TRUE}
# access the gain 
gain_ada <- gains(test_dataExpectancy$Life_expectancy, pred_ada) # creates the cumulative gain chart
print(gain_ada)

# Plot Lift chart: Percent cumulative response
x_ada <- c(0, gain_ada$depth)
pred.y_ada <- c(0, gain_ada$cume.pct.of.total)
avg.y_ada <- c(0, gain_ada$depth/100)
plot(x_ada, pred.y_ada, main = "Cumulative Lift Chart: Adaboost", xlab = "deciles", 
     ylab = "Percent cumulative response", type = "l", col = "blue", cex.lab = 1.5)
lines(x_ada, avg.y_ada, type = "l")
```

- Ridge Regression


```{r,echo=TRUE}
# access the gain 
gain_ridge <- gains(test_dataExpectancy$Life_expectancy, pred_ridge) # creates the cumulative gain chart
print(gain_ridge)

# Plot Lift chart: Percent cumulative response
x_ridge <- c(0, gain_ridge$depth)
pred.y_ridge <- c(0, gain_ridge$cume.pct.of.total)
avg.y_ridge <- c(0, gain_ridge$depth/100)
plot(x_ridge, pred.y_ridge, main = "Cumulative Lift Chart: Ridge Regression", xlab = "deciles", 
     ylab = "Percent cumulative response", type = "l", col = "blue", cex.lab = 1.5)
lines(x_ridge, avg.y_ridge, type = "l")
```




- Lasso Regression


```{r,echo=TRUE}
# access the gain 
gain_lasso <- gains(test_dataExpectancy$Life_expectancy, pred_lasso) # creates the cumulative gain chart
print(gain_lasso)

# Plot Lift chart: Percent cumulative response
x_lasso <- c(0, gain_lasso$depth)
pred.y_lasso <- c(0, gain_lasso$cume.pct.of.total)
avg.y_lasso <- c(0, gain_lasso$depth/100)
plot(x_lasso, pred.y_lasso, main = "Cumulative Lift Chart: Lasso regression", xlab = "deciles", 
     ylab = "Percent cumulative response", type = "l", col = "blue", cex.lab = 1.5)
lines(x_lasso, avg.y_lasso, type = "l")
```



- Knn algorithm lift plot 


```{r,echo=TRUE}
# access the gain 
gain_knn <- gains(test_dataExpectancy$Life_expectancy, pred_knn) # creates the cumulative gain chart
print(gain_knn)

# Plot Lift chart: Percent cumulative response
x_knn <- c(0, gain_knn$depth)
pred.y_knn <- c(0, gain_knn$cume.pct.of.total)
avg.y_knn <- c(0, gain_knn$depth/100)
plot(x_knn, pred.y_knn, main = "Cumulative Lift Chart: Lasso regression", xlab = "deciles", 
     ylab = "Percent cumulative response", type = "l", col = "blue", cex.lab = 1.5)
lines(x_knn, avg.y_knn, type = "l")
```


- tree regression 

```{r,echo=TRUE}
# access the gain 
gain_tree <- gains(test_dataExpectancy$Life_expectancy, pred_tree) # creates the cumulative gain chart
print(gain_tree)

# Plot Lift chart: Percent cumulative response
x_tree <- c(0, gain_tree$depth)
pred.y_tree <- c(0, gain_tree$cume.pct.of.total)
avg.y_tree <- c(0, gain_tree$depth/100)
plot(x_tree, pred.y_tree, main = "Cumulative Lift Chart: Decision tree", xlab = "deciles", 
     ylab = "Percent cumulative response", type = "l", col = "blue", cex.lab = 1.5)
lines(x_tree, avg.y_tree, type = "l")
```



- Decision tree 

```{r,echo=TRUE}
# plot the decision tree 
prp(tune_tree$finalModel, type = 1, extra = 1, under = TRUE, split.font = 2, varlen = -10)
```